A performance assessment of evolutionary algorithms in volunteer computing environments: the importance of entropy

### 2016-11-15 EvoPar2017/paper.Rnw
some corrections

The baseline, called {\em base} in the Table, uses the
mechanism of dropping contributions from clients if they are repeated
to avoid overlap. Clients still get a random chromosome from the
server, so they can in fact proceed with the algorithm and even finish
it. However, this was not known to the volunteer, so the fact that they were not contributing to the pool was
conveyed by inserting a graph in the client that represented the size
of the cache, labeled ``How am I contributing''. The user could
then realize there was no contribution and do something
about it. These sets of runs are labeled {\tt graph1} and 2. However, a stuck
client could still add to the pool if it found a new optimum, so we
added a mechanism for rebooting the client, that is, reloading the
page, if the cache size was found to be less than 1, which indicated a
run recently started. This batch of runs is labelled {\tt
  reboot}. This size of the cache was not really noticed by some slow clients or too fast runs, so that in the
last case, labeled {\tt cachecrash}, clients rebooted if the cache
collapsed by more than a certain length, indicating a cache that had
been voided and was filling again. Our objective was twofold: to
encourage engagement by the users and also to increase the involvement
of every client in the common experiment, eliminating at the same time
overlaps between runs.

%
\begin{table}
  \caption{Experiments with {\em overlapped} runs filtered out: summary of results. \label{tab:filtered}}
\begin{center}
\begin{tabular}{|l|c|c|c|c|c|c|c|}
\hline
Experiment & Median Time & \multicolumn{2}{|c|}{Median IPs} &\multicolumn{2}{|c|}{Median PUTs} & \#Runs &  Puts/IP \\
&   (ms)        & Total & Used & Total & Used &  & \\
\hline
base & 293685 & 13.0 & 8.0 & 483.0 & 57.0 & 47 & 35.50000\\
\hline
graph1 & 768053 & 12.0 & 5.0 & 601.0 & 79.0 & 21 & 45.84615\\
graph2 & 1477969 & 6.0 & 4.0 & 901.0 & 52.0 & 25 & 149.42857\\
\hline
reboot & 799471 & 9.5 & 6.5 & 656.5 & 67.0 & 44 & 76.48016\\
\hline
cachereboot & 463347 & 12.0 & 12.0 & 638.5 & 80.5 & 52 & 44.25911\\
\hline
\end{tabular}
%% Initially generated with this code, 
%%<<experiments.filtered, cache=FALSE,echo=FALSE>>=
%%experiment.aggregates.filtered$Experiment <- c("base","graph1","graph2","reboot","cachereboot")
%%kable(experiment.aggregates.filtered)
%%@ %def 
%
\end{center}
\end{table}
%
The first thing we will have to check is whether we have effectively eliminated these
overlaps. In order to do that, we will have to establish a threshold
under which we can reasonably expect that the solution cannot be found. In every
evolutionary algorithm it might happen, by chance, that the solution
is found in just a few evaluations, but that is usually not the
case. By looking at the logs in the last case we have established that
minimum as 16 PUTs; in less than 1600 generations no solution to that
difficult problem will be found. Let us filter then the runs,
eliminating those that used less than 1600 generations to find the
solution which can reasonably be said that they did so because some
browser ``carried over'' a chromosome from the previous run. The
results are shown in Table \ref{tab:filtered}. The first change is in
the column \#Runs, that shows that most of the runs in the initial
configurations found the solution thanks to these carried over
clients. However, in the last two rows, the ones that rebooted clients
when they detected changes in cache, the difference is minimal. This
means that one objective has been reached: if clients reboot when they
detect changes in the cache that imply a new run, overlaps are almost
eliminated. Going back to Table \ref{tab:runs}, we can see than the
ratio of effective contributions, those that feed the cache, increases
from around 13\% in the first case to almost 16\% in the last,
reaching lows of less than 6\% in some cases ({\tt graph2}). This
indicates that rebooting the clients has the positive effect of
increasing the contribution of clients to the common pool. However, we
should remark that the clients will still stop contributing if they
get stuck in a particular solution, so the ratio keeps being low
resulting in a slow change in the cache and a lower diversity of
clients that tap that cache. At any rate, an increased number of
effective PUTs will have a positive impact on the algorithm.

\begin{table}[htbp]
  \caption{Minute-wise correlation for number of participating volunteers. \label{tab:correlation}}
\begin{center}
<<correlations, cache=FALSE,echo=FALSE>>=
ip.correlations$Experiment <- c("base","graph1","reboot","cachereboot")
kable(ip.correlations)
@ %def correlations
\end{center}
\end{table}
%
We can check this impact by looking at the column that refers to
median time in Table \ref{tab:filtered} and its relation to the median
number of PUTs, which reflects the number of evaluations performed by
the clients. The last two rows have a low number of evaluations,
around 650, which is higher than the first two, probably due to the
lower volunteer supply, but better than the middle one; besides, the
amount is remarkably similar. But in the last case, it takes much less
time to find the solution {\em using the same number of
  evaluations}; the same situation applies if we compare {\tt graph1}
to {\tt cachereboot}. All variables are difficult to control in this
environment, but it might be due to the fact that parallelism has
increased, there are more volunteers contributing at the same time. In
fact, as it can be seen in the last column, the median number of
contributions per user is the lowest for {\tt cachereboot}, which
needs less contributions by users to
find the same solution. Besides, the last row shows that {\em all}
users actively contribute to the cache, while in the {\tt reboot}
strategy 30\% of users do not. This might be due to the fact that {\tt
  reboot} only does so when cache size is less than one, indicating
that this strategy is only partly successful in avoiding overlaps.

In fact, Table \ref{tab:correlation}, which shows
the correlation between the number of IPs from one minute to the
next, presents a similar  behavior for all experiments but the last one. A high correlation indicating that the number of volunteers changes
slowly from one minute to the next; however, the {\tt cachereboot} experiment correlation indicates a bigger
dynamism, with volunteers joining and leaving the experiment all the
time, maybe because experiments take less time and need less resources.

@@

The graph that plots the time to solution of every instance of the
experiment vs. entrop is shown in \ref{fig:entropy}. There is a clear
trend towards experiments with lower entropy taking more time and
experiments with higher entropy taking more time, which indeed can be
statistically fit to a linear model. This dependence on entropy is
much clearer than the assumed dependence on the number of users, which
is represented in the graph by dot color, and even the number of PUTs
or contributions, which is represented by dot size. You can see a
clearly blue and big dot, meaning more than 40 different IPs
participating and around 1500 PUTs in the high end of the graph,
showing that it took the longest to find the solution. On the other
hand, clearly red and small dots are on the lowest end of the graph,
indicating that it took a short time to find the solution as long as
the entropy is high.

>>>>

The baseline, called {\em base} in the Table, uses the
mechanism of dropping contributions from clients if they are repeated
to avoid overlap. Clients still get a random chromosome from the
server, so they can in fact proceed with the algorithm and even finish
it. However, this was not known to the volunteer, so the fact that they were not contributing to the pool was
conveyed by inserting a graph in the client that represented the size
of the cache, labeled ``How am I contributing''. The user could
then realize there was no contribution and do something
about it. These sets of runs are labeled {\tt graph1} and 2. However, a stuck
client could still add to the pool if it found a new optimum, so we
added a mechanism for rebooting the client, that is, reloading the
page, if the cache size was found to be less than 1, which indicated a
run recently started. This batch of runs is labelled {\tt
  reboot}. This size of the cache was not really noticed by some slow clients or too fast runs, so that in the
last case, is labeled {\tt cachecrash}, clients rebooted if the cache
collapsed by more than a certain length, indicating a cache that had
been voided and was filling again. Our objective was twofold: to
encourage user engagement and also to increase the involvement
of every client in the common experiment, eliminating at the same time
overlaps between runs.

%
\begin{table}
  \caption{Experiments with {\em overlapped} runs filtered out: summary of results. \label{tab:filtered}}
\begin{center}
\begin{tabular}{|l|c|c|c|c|c|c|c|}
\hline
Experiment & Median Time & \multicolumn{2}{|c|}{Median IPs} &\multicolumn{2}{|c|}{Median PUTs} & \#Runs &  Puts/IP \\
&   (ms)        & Total & Used & Total & Used &  & \\
\hline
base & 293685 & 13.0 & 8.0 & 483.0 & 57.0 & 47 & 35.50000\\
\hline
graph1 & 768053 & 12.0 & 5.0 & 601.0 & 79.0 & 21 & 45.84615\\
graph2 & 1477969 & 6.0 & 4.0 & 901.0 & 52.0 & 25 & 149.42857\\
\hline
reboot & 799471 & 9.5 & 6.5 & 656.5 & 67.0 & 44 & 76.48016\\
\hline
cachereboot & 463347 & 12.0 & 12.0 & 638.5 & 80.5 & 52 & 44.25911\\
\hline
\end{tabular}
%% Initially generated with this code, 
%%<<experiments.filtered, cache=FALSE,echo=FALSE>>=
%%experiment.aggregates.filtered$Experiment <- c("base","graph1","graph2","reboot","cachereboot")
%%kable(experiment.aggregates.filtered)
%%@ %def 
%
\end{center}
\end{table}
%
The first thing we will have to check is whether we have effectively eliminated these
overlaps. In order to do that, we will have to establish a threshold
under which we can reasonably expect that the solution cannot be found. In every
evolutionary algorithm it might happen, by chance, that the solution
is found in just a few evaluations, but that is usually not the
case. By looking at the logs in the last case we have established that
minimum as 16 PUTs; in less than 1600 generations no solution to that
difficult problem will be found. Let us filter then the runs,
eliminating those that used less than 1600 generations to find the
solution which can reasonably be said that they did so because some
browser ``carried over'' a chromosome from the previous run. The
results are shown in Table \ref{tab:filtered}. The first change is in
the column \#Runs, that shows that most of the runs in the initial
configurations found the solution thanks to these carried over
clients. However, in the last two rows, the ones that rebooted clients
when they detected changes in cache, the difference is minimal. This
means that one objective has been reached: if clients reboot when they
detect changes in the cache implying a new run, this way overlaps are almost
eliminated. Going back to Table \ref{tab:runs}, we can see than the
ratio of effective contributions, those that feed the cache, increases
from around 13\% in the first case to almost 16\% in the last,
reaching lows of less than 6\% in some cases ({\tt graph2}). This
indicates that rebooting the clients has the positive effect of
increasing the contribution of clients to the common pool. However, we
should remark that some clients will still stop contributing if they
get stuck in a particular solution, so the ratio keeps being low
resulting in a slow change in the cache and a lower diversity of
clients that tap that cache. At any rate, an increased number of
effective PUTs will have a positive impact on the algorithm.

\begin{table}[htbp]
  \caption{Minute-wise correlation for number of participating volunteers. \label{tab:correlation}}
\begin{center}
<<correlations, cache=FALSE,echo=FALSE>>=
ip.correlations$Experiment <- c("base","graph1","reboot","cachereboot")
kable(ip.correlations)
@ %def correlations
\end{center}
\end{table}
%
We can check this impact by looking at the column that refers to
median time in Table \ref{tab:filtered} and its relation to the median
number of PUTs, which reflects the number of evaluations performed by
the clients. The last two rows have a low number of evaluations,
around 650, which is higher than the first two, probably due to the
lower volunteer supply, but better than the middle one; besides, the
amount is remarkably similar. But in the last case, it takes much less
time to find the solution {\em using the same number of
  evaluations}; the same situation applies if we compare {\tt graph1}
to {\tt cachereboot}. All variables are difficult to control in this
environment, but it might be due to the fact that parallelism has
increased, there are more volunteers contributing at the same time. In
fact, as it can be seen in the last column, the median number of
contributions per user is the lowest for {\tt cachereboot}, which
needs less contributions by users to
find the same solution. Besides, the last row shows that {\em all}
users actively contribute to the cache, while in the {\tt reboot}
strategy 30\% of users do not. This might be due to the fact that {\tt
  reboot} is done only when the cache size is less than one, indicating
that this strategy is only partly successful in avoiding overlaps.

In fact, Table \ref{tab:correlation}, shows
the correlation between the number of IPs from one minute to the
next, presents a similar  behavior for all experiments but the last one. A high correlation indicating that the number of volunteers changes
slowly from one minute to the next; however, the {\tt cachereboot} experiment correlation indicates a bigger
dynamism, with volunteers joining and leaving the experiment all the
time, maybe because experiments take less time and need less resources.

@@

The graph that plots the time to solution of every instance of the
experiment vs. entropy is shown in \ref{fig:entropy}. There is a clear
trend towards experiments with lower entropy taking more time and
experiments with higher entropy taking more time, which indeed can be
statistically fit to a linear model. This dependence on entropy is
much clearer than the assumed dependence on the number of users, which
is represented in the graph by dot color, and even the number of PUTs
or contributions, which is represented by dot size. You can see a
clearly blue and big dot, meaning more than 40 different IPs
participating and around 1500 PUTs in the high end of the graph,
showing that it took the longest to find the solution. On the other
hand, clearly red and small dots are on the lowest end of the graph,
indicating that it took a short time to find the solution as long as
the entropy is high.
